paper figures

### 2023-09-08 scRNAseq/paper_figures.Rmd
update figures

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE, dev=c('png','pdf'))
library(sctransform)
library(Seurat)
library(SeuratDisk)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(dplyr)
library(scRepertoire)
library(tidyr)
library(cluster)
library(plyr)
library(pheatmap)
library(dendextend)
library(ggpubr)
library(ComplexHeatmap)
library(RColorBrewer)
library(Binarize)
options(future.globals.maxSize = 10 * 1024 * 1024^2)
path <- getwd()
```

@@

```{r normalize}
for (nn in names(pbmc)) {
  DefaultAssay(pbmc[[nn]]) <- 'RNA'
  pbmc[[nn]] <- NormalizeData(pbmc[[nn]],normalization.method='LogNormalize',scale.factor=10000,verbose=FALSE)
}
```

# Fig 1 + supplements

@@

```{r Fig_1C,fig.width=7,fig.height=5}
ct_map <- c('CD8 TCM'='Mmry CD8+ T cells',
            'CD8 TEM'='Mmry CD8+ T cells',
            'CD8 Naive'='Naive CD8+ T cells', 
            'CD4 TCM'='Mmry CD4+ T cells',
            'CD4 TEM'='Mmry CD4+ T cells',
            'CD4 Naive'='Naive CD4+ T cells',
            'NK'='NK cells',
            'NK Proliferating'='NK cells',
            'NK_CD56bright'='NK cells',
            'gdT'='gd T cells',
            'MAIT'='MAIT cells',
            'B intermediate'='IgD+ B cells',
            'B naive'='IgD+ B cells',
            'B memory'='IgD- B cells',
            'CD14 Mono'='cMonocytes',
            'CD16 Mono'='ncMonocytes',
            'CD4 CTL'='Cytotoxic T cells',
            'CD4 Proliferating'='Mmry CD4+ T cells',
            'CD8 Proliferating'='Mmry CD8+ T cells',
            'cDC1'='cDCs',
            'cDC2'='cDCs',
            'pDC'='pDCs',
            'Doublet'='other',
            'Eryth'='other',
            'dnT'='other',
            'ASDC'='other',
            'HSPC'='other',
            'ILC'='ILCs',
            'Plasmablast'='Plasmablast',
            'Platelet'='other',
            'Treg'='Tregs')

@@

```{r Fig_S3B,fig.width=4.5,fig.height=3.5}
ggplot(pbmc_meta %>% dplyr::sample_frac(1L), 
       aes(x=refUMAP_1,y=refUMAP_2,color=Patient.ID)) + 
  geom_point(size=.25) +
  theme_void() + 
  coord_fixed() +
  guides(color=guide_legend(override.aes = list(size=3))) +
  theme(legend.key.size=unit(3.5,'mm'))
```

@@

```{r Fig_S3A,fig.width=7,fig.height=4.5, cache=FALSE}
ABs <- c("CD3","CD8a","CD40","CD19","CD20","CD45RA","CD38","CD69","HLA-DR",
         "CD11c","CD123","CD16","CD161","CD1c","CD14","CD25","CD127","CD56","CD4","CD45")

@@

```{r Fig_S3C, fig.width=7,fig.height=4}
#library(lme4)
library(gtools)

@@

```{r Fig_1E,fig.width=4.5,fig.height=4, cache=FALSE}
for (nn in grep('PBMC',names(pbmc),value=TRUE)) {
  DefaultAssay(pbmc[[nn]]) <- 'RNA'
}

@@

```{r Fig_S8A,fig.width=7,fig.height=3}
RNA %>% 
  spread(gene,expression) %>% 
  dplyr::group_by(orig.ident,Patient.ID, celltype,Timepoint,Diagnosis) %>% 
  dplyr::summarise(pos=mean((ITGA4 > 0) & (ITGB7 > 0))) %>%
  dplyr::filter(Timepoint %in% c("Vedo_0d","ctrl")) %>%
  dplyr::mutate(condition=factor(ifelse(Diagnosis=='control','HC','UC'),levels=c('HC','UC'))) %>%
  ggplot(aes(x=condition,y=pos,fill=condition,color=condition)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('UC','HC')),label = "p.format", 
                     method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  facet_grid(.~celltype,switch='both') +
  scale_fill_manual(values=condition_colors,limits=force) +
  scale_color_manual(values=condition_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='fraction ITGA4/ITGB7\ndouble-positive cells',fill='',color='')
```

# Fig 2

@@

```{r Fig_2A,fig.width=5,fig.height=4}
RNA <- do.call(rbind,
               lapply(pbmc[grepl('PBMC',names(pbmc))], 
                      function(sobj) FetchData(sobj, c('ITGA4','ITGB7','orig.ident')) %>%
                        tibble::rownames_to_column('cell') %>%
                        dplyr::mutate(cell=paste0(orig.ident,'_',cell)) %>%
                        dplyr::select(-orig.ident))) %>%
  dplyr::left_join(pbmc_meta,by='cell') %>%
  gather(gene,expression,c(ITGA4,ITGB7)) %>%
  dplyr::arrange(gene,expression)

@@

```{r Fig_2B,fig.width=3.75,fig.height=4, cache=FALSE}
for (nn in grep('PBMC',names(pbmc),value=TRUE)) {
  DefaultAssay(pbmc[[nn]]) <- 'orig_ADT'
}

@@

# Fig. 3 + supplements

@@

```{r Fig_3B,fig.width=4.5,fig.height=3.5}
pbmc.T <- lapply(names(pbmc), function(sample) {
  bcs <- tcr_combined[[sample]]$barcode
  cells <- paste0(sample,'_',sample,'_',Cells(pbmc[[sample]]))
  tmp <- RenameCells(pbmc[[sample]], new.names=cells)
  combineExpression(tcr_combined[[sample]][bcs %in% cells,],tmp, cloneCall="aa")
  })
names(pbmc.T) <- names(pbmc)

@@

```{r clonotype_celltypes_TCR,fig.width=6.5,fig.height=10}
tmp <- list()
for (sample in grep('PBMC',names(pbmc.T),value=TRUE)) {
  sobj <- pbmc.T[[sample]]
  sobj$celltype <- plyr::revalue(sobj$predicted.celltype.l2,ct_map)
  sobj$cloneType <- factor(tolower(gsub('Single','Small',gsub(' .*','',sobj$cloneType))),
                           levels=c('small','medium','large','hyperexpanded'))
  tmp[[sample]] <- occupiedscRepertoire(sobj,
                                        x.axis = "celltype", 
                                        exportTable=TRUE) %>%
    dplyr::mutate(sample=gsub('_PBMC','',sample))
}

ggplot(do.call(rbind,tmp),aes(x=celltype,y=value,fill=cloneType)) + 
  geom_col() + 
  theme_classic() +
  facet_wrap(~sample,scales='free_y',ncol=2,strip.position='left') +
  scale_fill_manual(values=cloneType_colors) +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5),
        strip.background=element_blank(),
        strip.placement='outside',
        legend.key.size=unit(3,'mm')) +
  labs(x='',y='')
```

@@

```{r Fig_3C,fig.width=6,fig.height=2}
cols <- c('sample','orig.ident','predicted.celltype.l2')
all_meta <- do.call(rbind,lapply(pbmc[grepl('PBMC',names(pbmc))], function(x) x@meta.data[,cols])) %>%
  tibble::rownames_to_column('cell') %>%
  dplyr::mutate(celltype=plyr::revalue(predicted.celltype.l2,ct_map),
                cell=paste0(sample,'_',gsub('\\.','_',cell))) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=ifelse(Characteristics.Responder.=='yes','R','NR'),
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl')),
                                   Material=ifelse(Characteristics.Material.=='PBMC','PBMC','CD4')) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Material,Timepoint,Batch),
                   by=c('sample'='Sample.Name'))

dfs <- list()
for (ct in grep('CD[48]|Treg',unique(all_meta$celltype),value=TRUE)) {
  bcs <- all_meta$cell[all_meta$celltype==ct]
  tcrcomb <- lapply(tcr_combined,
                    function(x) x[(x$barcode %in% bcs) & !is.na(x$TCR1) & !is.na(x$TCR2),])
  dfs[[ct]] <- clonalDiversity(tcrcomb[lapply(tcrcomb,function(x) dim(x)[1]) > 0],
                               cloneCall='gene',group='sample',exportTable=TRUE, skip.boots=TRUE) %>%
    dplyr::mutate(celltype=ct)
}
df <- do.call(rbind,dfs) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=ifelse(Characteristics.Responder.=='yes','R','NR'),
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl'))) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Timepoint,Batch),
                   by=c('sample'='Sample.Name')) %>%
  dplyr::arrange(Patient.ID)

@@

```{r Fig_S11F,fig.width=6,fig.height=3}
pl <- ggplot(df %>% 
               dplyr::filter(Timepoint!='ctrl',celltype!='Tregs') %>%
               gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype)) %>%
               dplyr::filter(metric=='Inv.Simpson') %>%
               dplyr::arrange(Responder,celltype,Timepoint,Patient.ID),
             aes(x=Responder,y=value)) +
  geom_boxplot(aes(fill=Responder,color=Responder),outlier.shape=NA) +
  geom_point(size=1,color='gray30') +
  theme_classic() +
  coord_cartesian(clip='off') +
  scale_y_log10() +
  facet_grid(Timepoint ~ celltype,scales='free_y',switch='y') +
  stat_compare_means(comparisons=list(c('NR','R')),label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
  scale_fill_manual(values=responder_colors,limits=force) +
  scale_color_manual(values=responder_colors,limits=force) +
  labs(x='',y='Inv. Simpson',title='') +
  theme(legend.key.size=unit(4,'mm'),
        legend.box='horizontal',
        axis.text.x=element_blank(),
        strip.background=element_blank(),
        strip.text=element_text(size=8),
        strip.placement='outside')

@@

```{r Figs_3E_3F_S11B_S11C_S11D,fig.width=2.5,fig.height=2.5, cache=FALSE}
library(Binarize)
for (nn in grep('PBMC',names(pbmc),value=TRUE)) {
  DefaultAssay(pbmc[[nn]]) <- 'RNA'
}

@@

for (ss in c('CCR9','GPR15','Integrin_b7','ITGA4','ITGAE','ITGB7')) {
  for (ct in c('Mmry CD4','Mmry CD8')) {
    pl <- ggplot(df %>% 
                   dplyr::filter(Diagnosis=='UC') %>%
                   gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype,subset)) %>%
                   dplyr::filter(metric=='Inv.Simpson') %>%
                   dplyr::group_by(celltype,subset,Patient.ID) %>%
                   dplyr::mutate(ns=n()) %>%
                   dplyr::filter(ns==2,subset==ss,celltype==ct), 
                 aes(x=Timepoint,y=value)) +
      geom_boxplot(aes(fill=Timepoint,color=Timepoint),outlier.shape=NA) +
      geom_line(aes(group=Patient.ID),color='gray',size=.25) +
      geom_point(size=1,color='gray30') +
      theme_classic() +
      coord_cartesian(clip='off') +
      scale_y_log10() +
      stat_compare_means(comparisons=list(c('Vedo_6wk','Vedo_0d')),paired=TRUE,label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
      scale_fill_manual(values=timepoint_colors) +
      scale_color_manual(values=timepoint_colors) +
      labs(x='',y='Inv. Simpson',title=paste0(ct,'\n',ss)) +
      theme(legend.key.size=unit(4,'mm'),
            legend.box='horizontal',
            axis.text.x=element_text(angle=90,hjust=1,vjust=0),
            strip.background=element_blank(),
            strip.placement='outside')
    print(pl)
  }
}
```

```{r Fig_3H,fig.width=2.5,fig.height=2}
ADT %>% 
  dplyr::group_by(orig.ident,Patient.ID, celltype,Timepoint) %>% 
  dplyr::summarise(pos=mean((expression > threshold))) %>%
  dplyr::filter(Timepoint %in% c("Vedo_0d","Vedo_6wk")) %>%
  dplyr::group_by(celltype,Patient.ID) %>%
  dplyr::mutate(ns=n()) %>%
  dplyr::filter(ns==2, celltype=='Mmry CD4+ T cells') %>%
  ggplot(aes(x=Timepoint,y=pos,fill=Timepoint,color=Timepoint)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  geom_line(aes(group=Patient.ID),size=.25,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('Vedo_6wk','Vedo_0d')),label = "p.format", 
                     paired=TRUE,method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  coord_cartesian(clip='off') +
  #facet_grid(.~celltype,switch='both') +
  scale_fill_manual(values=timepoint_colors,limits=force) +
  scale_color_manual(values=timepoint_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=90,hjust=1,vjust=.5))  +
  labs(x='',y='fraction Integrin-b7-positive cells',fill='',color='')
```

```{r Fig_3G, fig.width=2.5,fig.height=2.}
pbmc_frac %>%
  dplyr::filter(Timepoint %in% c('Vedo_0d','Vedo_6wk')) %>%
  dplyr::group_by(celltype,Patient.ID) %>%
  dplyr::mutate(ns=n()) %>%
  dplyr::filter(ns==2,celltype=='Mmry CD4+ T cells') %>%
  ggplot(aes(x=Timepoint,y=frac,fill=Timepoint,color=Timepoint)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  geom_line(aes(group=Patient.ID),size=.25,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('Vedo_6wk','Vedo_0d')),label = "p.format", 
                     paired=TRUE,method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  #facet_grid(.~celltype) +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=timepoint_colors,limits=force) +
  scale_color_manual(values=timepoint_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='cell fraction',fill='',color='',title='Mmry CD4+ T cells')
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE, dev=c('png','pdf'))
library(sctransform)
library(Seurat)
library(SeuratDisk)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(dplyr)
library(scRepertoire)
library(tidyr)
library(cluster)
library(plyr)
library(pheatmap)
library(dendextend)
library(ggpubr)
library(ComplexHeatmap)
library(RColorBrewer)
library(Binarize)
library(limma)
options(future.globals.maxSize = 10 * 1024 * 1024^2)
path <- getwd()
```

@@

```{r normalize, cache=FALSE}
for (nn in names(pbmc)) {
  DefaultAssay(pbmc[[nn]]) <- 'RNA'
  pbmc[[nn]] <- NormalizeData(pbmc[[nn]],normalization.method='LogNormalize',scale.factor=10000,verbose=FALSE)
}
```

@@

```{r Fig_S2A,fig.width=4.5,fig.height=3.5}
ct_map <- c('CD8 TCM'='Mmry CD8+ T cells',
            'CD8 TEM'='Mmry CD8+ T cells',
            'CD8 Naive'='Naive CD8+ T cells', 
            'CD4 TCM'='Mmry CD4+ T cells',
            'CD4 TEM'='Mmry CD4+ T cells',
            'CD4 Naive'='Naive CD4+ T cells',
            'NK'='NK cells',
            'NK Proliferating'='NK cells',
            'NK_CD56bright'='NK cells',
            'gdT'='gd T cells',
            'MAIT'='MAIT cells',
            'B intermediate'='IgD+ B cells',
            'B naive'='IgD+ B cells',
            'B memory'='IgD- B cells',
            'CD14 Mono'='cMonocytes',
            'CD16 Mono'='ncMonocytes',
            'CD4 CTL'='Cytotoxic T cells',
            'CD4 Proliferating'='Mmry CD4+ T cells',
            'CD8 Proliferating'='Mmry CD8+ T cells',
            'cDC1'='cDCs',
            'cDC2'='cDCs',
            'pDC'='pDCs',
            'Doublet'='other',
            'Eryth'='other',
            'dnT'='other',
            'ASDC'='other',
            'HSPC'='other',
            'ILC'='ILCs',
            'Plasmablast'='Plasmablast',
            'Platelet'='other',
            'Treg'='Tregs')

@@

```{r Fig_S2B,fig.width=4.5,fig.height=3.5}
ggplot(pbmc_meta %>% dplyr::sample_frac(1L), 
       aes(x=refUMAP_1,y=refUMAP_2,color=Patient.ID)) + 
  geom_point(size=.25) +
  theme_void() + 
  coord_fixed() +
  guides(color=guide_legend(override.aes = list(size=3))) +
  theme(legend.key.size=unit(3.5,'mm'))
```

@@

```{r Fig_S2C,fig.width=7,fig.height=4.5, cache=FALSE}
ABs <- c("CD3","CD8a","CD40","CD19","CD20","CD45RA","CD38","CD69","HLA-DR",
         "CD11c","CD123","CD16","CD161","CD1c","CD14","CD25","CD127","CD56","CD4","CD45")

@@

```{r Fig_S2D, fig.width=7,fig.height=4}
#library(lme4)
library(gtools)

@@

```{r Fig_S3A,fig.width=4.5,fig.height=4, cache=FALSE}
for (nn in grep('PBMC',names(pbmc),value=TRUE)) {
  DefaultAssay(pbmc[[nn]]) <- 'RNA'
}

@@

```{r Fig_S3D,fig.width=7,fig.height=3}
RNA %>% 
  spread(gene,expression) %>% 
  dplyr::group_by(orig.ident,Patient.ID, celltype,Timepoint,Diagnosis) %>% 
  dplyr::summarise(pos=mean((ITGA4 > 0) & (ITGB7 > 0))) %>%
  dplyr::filter(Timepoint %in% c("Vedo_0d","ctrl")) %>%
  dplyr::mutate(condition=factor(ifelse(Diagnosis=='control','HC','UC'),levels=c('HC','UC'))) %>%
  ggplot(aes(x=condition,y=pos,fill=condition,color=condition)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('UC','HC')),label = "p.format", 
                     method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  facet_grid(.~celltype,switch='both') +
  scale_fill_manual(values=condition_colors,limits=force) +
  scale_color_manual(values=condition_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='fraction ITGA4/ITGB7\ndouble-positive cells',fill='',color='')
```

@@

```{r Fig_S3B,fig.width=5,fig.height=4}
RNA <- do.call(rbind,
               lapply(pbmc[grepl('PBMC',names(pbmc))], 
                      function(sobj) FetchData(sobj, c('ITGA4','ITGB7','orig.ident')) %>%
                        tibble::rownames_to_column('cell') %>%
                        dplyr::mutate(cell=paste0(orig.ident,'_',cell)) %>%
                        dplyr::select(-orig.ident))) %>%
  dplyr::left_join(pbmc_meta,by='cell') %>%
  gather(gene,expression,c(ITGA4,ITGB7)) %>%
  dplyr::arrange(gene,expression)

@@

```{r Fig_S3C,fig.width=3.75,fig.height=4, cache=FALSE}
for (nn in grep('PBMC',names(pbmc),value=TRUE)) {
  DefaultAssay(pbmc[[nn]]) <- 'orig_ADT'
}

@@

@@

```{r Fig_S4B,fig.width=4.5,fig.height=3.5, cache=FALSE}
pbmc.T <- lapply(names(pbmc), function(sample) {
  bcs <- tcr_combined[[sample]]$barcode
  cells <- paste0(sample,'_',sample,'_',Cells(pbmc[[sample]]))
  tmp <- RenameCells(pbmc[[sample]], new.names=cells)
  combineExpression(tcr_combined[[sample]][bcs %in% cells,],tmp, cloneCall="aa")
  })
names(pbmc.T) <- names(pbmc)

@@

<!-- ```{r clonotype_celltypes_TCR,fig.width=6.5,fig.height=10} -->
<!-- tmp <- list() -->
<!-- for (sample in grep('PBMC',names(pbmc.T),value=TRUE)) { -->
<!--   sobj <- pbmc.T[[sample]] -->
<!--   sobj$celltype <- plyr::revalue(sobj$predicted.celltype.l2,ct_map) -->
<!--   sobj$cloneType <- factor(tolower(gsub('Single','Small',gsub(' .*','',sobj$cloneType))), -->
<!--                            levels=c('small','medium','large','hyperexpanded')) -->
<!--   tmp[[sample]] <- occupiedscRepertoire(sobj, -->
<!--                                         x.axis = "celltype",  -->
<!--                                         exportTable=TRUE) %>% -->
<!--     dplyr::mutate(sample=gsub('_PBMC','',sample)) -->
<!-- } -->

<!-- ggplot(do.call(rbind,tmp),aes(x=celltype,y=value,fill=cloneType)) +  -->
<!--   geom_col() +  -->
<!--   theme_classic() + -->
<!--   facet_wrap(~sample,scales='free_y',ncol=2,strip.position='left') + -->
<!--   scale_fill_manual(values=cloneType_colors) + -->
<!--   theme(axis.text.x=element_text(angle=90,hjust=1,vjust=.5), -->
<!--         strip.background=element_blank(), -->
<!--         strip.placement='outside', -->
<!--         legend.key.size=unit(3,'mm')) + -->
<!--   labs(x='',y='') -->
<!-- ``` -->

@@

```{r Fig_S4C,fig.width=6,fig.height=2}
cols <- c('sample','orig.ident','predicted.celltype.l2')
all_meta <- do.call(rbind,lapply(pbmc[grepl('PBMC',names(pbmc))], function(x) x@meta.data[,cols])) %>%
  tibble::rownames_to_column('cell') %>%
  dplyr::mutate(celltype=plyr::revalue(predicted.celltype.l2,ct_map),
                cell=paste0(sample,'_',gsub('\\.','_',cell))) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=ifelse(Characteristics.Responder.=='yes','R','NR'),
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint=ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl')),
                                   Material=ifelse(Characteristics.Material.=='PBMC','PBMC','CD4')) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Material,Timepoint,Batch),
                   by=c('sample'='Sample.Name'))

dfs <- list()
for (ct in grep('CD[48]|Treg',unique(all_meta$celltype),value=TRUE)) {
  bcs <- all_meta$cell[all_meta$celltype==ct]
  tcrcomb <- lapply(tcr_combined,
                    function(x) x[(x$barcode %in% bcs) & !is.na(x$TCR1) & !is.na(x$TCR2),])
  dfs[[ct]] <- clonalDiversity(tcrcomb[lapply(tcrcomb,function(x) dim(x)[1]) > 0],
                               cloneCall='gene',group='sample',exportTable=TRUE, skip.boots=TRUE) %>%
    dplyr::mutate(celltype=ct)
}
df <- do.call(rbind,dfs) %>%
  dplyr::left_join(study %>% 
                     dplyr::mutate(Patient.ID=gsub('[P]*([0-9]*)[\\/_][12]','P\\1',Characteristics.Patient.ID.),
                                   Responder=ifelse(Characteristics.Responder.=='yes','R','NR'),
                                   Diagnosis=Characteristics.Diagnosis.,
                                   Batch=gsub(' ','',gsub('_scRNAseq','',Parameter.Value.Method.)),
                                   Timepoint =ifelse(grepl('_1_',Sample.Name),'Vedo_0d',
                                                    ifelse(grepl('_2_',Sample.Name),'Vedo_6wk','ctrl'))) %>%
                     dplyr::select(Patient.ID,Responder,Diagnosis,Sample.Name,Timepoint,Batch),
                   by=c('sample'='Sample.Name')) %>%
  dplyr::arrange(Patient.ID)

@@

<!-- ```{r Fig_S11F,fig.width=6,fig.height=3} -->
<!-- pl <- ggplot(df %>%  -->
<!--                dplyr::filter(Timepoint!='ctrl',celltype!='Tregs') %>% -->
<!--                gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype)) %>% -->
<!--                dplyr::filter(metric=='Inv.Simpson') %>% -->
<!--                dplyr::arrange(Responder,celltype,Timepoint,Patient.ID), -->
<!--              aes(x=Responder,y=value)) + -->
<!--   geom_boxplot(aes(fill=Responder,color=Responder),outlier.shape=NA) + -->
<!--   geom_point(size=1,color='gray30') + -->
<!--   theme_classic() + -->
<!--   coord_cartesian(clip='off') + -->
<!--   scale_y_log10() + -->
<!--   facet_grid(Timepoint ~ celltype,scales='free_y',switch='y') + -->
<!--   stat_compare_means(comparisons=list(c('NR','R')),label = "p.format",size=2,bracket.size=.25,tip.length=.01) + -->
<!--   scale_fill_manual(values=responder_colors,limits=force) + -->
<!--   scale_color_manual(values=responder_colors,limits=force) + -->
<!--   labs(x='',y='Inv. Simpson',title='') + -->
<!--   theme(legend.key.size=unit(4,'mm'), -->
<!--         legend.box='horizontal', -->
<!--         axis.text.x=element_blank(), -->
<!--         strip.background=element_blank(), -->
<!--         strip.text=element_text(size=8), -->
<!--         strip.placement='outside') -->

<!-- pg <- ggplotGrob(pl) -->
<!-- for(i in which(grepl("strip-t", pg$layout$name))){ -->
<!--   pg$grobs[[i]]$layout$clip <- "off" -->
<!-- } -->

<!-- grid::grid.draw(pg) -->
<!-- #pl -->
<!-- ``` -->

@@

```{r Fig_S4E_S4H_S5C_S5D,fig.width=2.5,fig.height=2.5, cache=FALSE}
library(Binarize)
for (nn in grep('PBMC',names(pbmc),value=TRUE)) {
  DefaultAssay(pbmc[[nn]]) <- 'RNA'
}

@@

for (ss in c('GPR15','Integrin_b7','ITGA4','ITGAE','ITGB7')) {
  for (ct in c('Mmry CD4','Mmry CD8')) {
    pl <- ggplot(df %>% 
                   dplyr::filter(Diagnosis=='UC') %>%
                   gather(metric,value,-c(sample,Patient.ID,Responder,Diagnosis,Timepoint,Batch,celltype,subset)) %>%
                   dplyr::filter(metric=='Inv.Simpson') %>%
                   dplyr::group_by(celltype,subset,Patient.ID) %>%
                   dplyr::mutate(ns=n()) %>%
                   dplyr::filter(ns==2,subset==ss,celltype==ct), 
                 aes(x=Timepoint,y=value)) +
      geom_boxplot(aes(fill=Timepoint,color=Timepoint),outlier.shape=NA) +
      geom_line(aes(group=Patient.ID),color='gray',size=.25) +
      geom_point(size=1,color='gray30') +
      theme_classic() +
      coord_cartesian(clip='off') +
      scale_y_log10() +
      stat_compare_means(comparisons=list(c('Vedo_6wk','Vedo_0d')),paired=TRUE,label = "p.format",size=2,bracket.size=.25,tip.length=.01) +
      scale_fill_manual(values=timepoint_colors) +
      scale_color_manual(values=timepoint_colors) +
      labs(x='',y='Inv. Simpson',title=paste0(ct,'\n',ss)) +
      theme(legend.key.size=unit(4,'mm'),
            legend.box='horizontal',
            axis.text.x=element_text(angle=90,hjust=1,vjust=0),
            strip.background=element_blank(),
            strip.placement='outside')
    print(pl)
  }
}
```

```{r Fig_S4H,fig.width=2.5,fig.height=2}
ADT %>% 
  dplyr::group_by(orig.ident,Patient.ID, celltype,Timepoint) %>% 
  dplyr::summarise(pos=mean((expression > threshold))) %>%
  dplyr::filter(Timepoint %in% c("Vedo_0d","Vedo_6wk")) %>%
  dplyr::group_by(celltype,Patient.ID) %>%
  dplyr::mutate(ns=n()) %>%
  dplyr::filter(ns==2, celltype=='Mmry CD4+ T cells') %>%
  ggplot(aes(x=Timepoint,y=pos,fill=Timepoint,color=Timepoint)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  geom_line(aes(group=Patient.ID),size=.25,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('Vedo_6wk','Vedo_0d')),label = "p.format", 
                     paired=TRUE,method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  coord_cartesian(clip='off') +
  #facet_grid(.~celltype,switch='both') +
  scale_fill_manual(values=timepoint_colors,limits=force) +
  scale_color_manual(values=timepoint_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=90,hjust=1,vjust=.5))  +
  labs(x='',y='fraction Integrin-b7-positive cells',fill='',color='')
```

```{r Fig_S4G, fig.width=2.5,fig.height=2.}
pbmc_frac %>%
  dplyr::filter(Timepoint %in% c('Vedo_0d','Vedo_6wk')) %>%
  dplyr::group_by(celltype,Patient.ID) %>%
  dplyr::mutate(ns=n()) %>%
  dplyr::filter(ns==2,celltype=='Mmry CD4+ T cells') %>%
  ggplot(aes(x=Timepoint,y=frac,fill=Timepoint,color=Timepoint)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  geom_line(aes(group=Patient.ID),size=.25,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('Vedo_6wk','Vedo_0d')),label = "p.format", 
                     paired=TRUE,method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  #facet_grid(.~celltype) +
  coord_cartesian(clip='off') +
  scale_fill_manual(values=timepoint_colors,limits=force) +
  scale_color_manual(values=timepoint_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='cell fraction',fill='',color='',title='Mmry CD4+ T cells')
```

```{r Fig_S5E, fig.width=2.5,fig.height=2.}
pbmc_frac %>%
  dplyr::filter(Timepoint %in% c('Vedo_0d','Vedo_6wk')) %>%
  dplyr::group_by(celltype,Patient.ID) %>%
  dplyr::mutate(ns=n()) %>%
  dplyr::filter(ns==2,celltype=='Mmry CD8+ T cells') %>%
  ggplot(aes(x=Timepoint,y=frac,fill=Timepoint,color=Timepoint)) + 
  geom_boxplot(outlier.shape=NA) +
  geom_point(size=.5,color='gray30') +
  geom_line(aes(group=Patient.ID),size=.25,color='gray30') +
  stat_compare_means(aes(group=celltype),comparisons=list(c('Vedo_6wk','Vedo_0d')),label = "p.format", 
                     paired=TRUE,method='wilcox.test',bracket.size=.25,tip.length=.01,size=2) +
  theme_classic() + 
  coord_cartesian(clip='off') +
  scale_fill_manual(values=timepoint_colors,limits=force) +
  scale_color_manual(values=timepoint_colors,limits=force) +
  theme(axis.text.x=element_blank(),
        legend.position='right',
        legend.key.size=unit(3.5,'mm'),
        strip.background=element_blank(),
        strip.placement='outside',
        strip.text=element_text(size=8,angle=45,hjust=1,vjust=1))  +
  labs(x='',y='cell fraction',fill='',color='',title='Mmry CD8+ T cells')
```
